Numerical analysis of the linear vertical shear instability in astrophysical disks

### 2014-12-27 linear_adia.tex
extended analytic discussion to thermal relaxation cases

\subsubsection{Explicit solutions in the thin-disk limit}
We can construct solutions in the thin-disk limit, in which 
$\ln(\rho/\rho_0) \simeq -\zhat^2/2$. Eq. \ref{adia_iso3} becomes   
\begin{align}
  0 = \delta v_z^{\prime\prime} - \zhat A \delta v_z^\prime + \left(B
    - C \zhat^2\right)\delta v_z,
\end{align}
with
\begin{align}
  &A \equiv 1 + \ii \epsilon q \hat{k},\\
  &B \equiv \hat{\sigma}^2\left(\frac{1}{\gamma} + \hat{k}^2\right) -
  \left(\frac{1}{\gamma} + \ii \hat{k} \epsilon q\right)\\
  &C \equiv \frac{\left(\gamma-1\right)}{\gamma}\hat{k}^2\left(1 - \frac{\ii
      \epsilon q}{\hat{k}}\right).\label{adia_thin}
\end{align}
Let
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\alpha\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where $\alpha$ is a constant to be chosen for convenience. Inserting
Eq. \ref{adia_ansatz} into Eq. \ref{adia_thin} gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime + \left(B +
    \alpha\right)g
  +\left(\alpha^2 - \alpha A - C\right)\zhat^2 g.
\end{align}
We choose $\alpha$ to make the coefficient of $\zhat^2g$
vanish, and impose the vertical kinetic energy density
$\rho|\delta v_z|^2$ to remain finite. Then assuming $g(\zhat)$ is a
polynomial, we require  
\begin{align}
  \real\alpha < \frac{1}{2}. 
\end{align}
We therefore choose 
\begin{align}
  \alpha = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right). 
\end{align}

We now have an equation in the same form as Eq. \ref{iso_ode3}:
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
% We again seek power-series solutions 
% \begin{align}
%   g(\zhat) = \sum_{m=0}^\infty b_m \zhat^m,
% \end{align}
% which requires
% \begin{align}
%   (m+2)(m+1)b_{m+2} + \left[\left(2\alpha - A\right)m + B + \alpha\right]=0.
% \end{align}
which requires
\begin{align}
  B + \alpha = M\left(A-2\alpha\right).
\end{align}
Inserting the definition of $B$ we get
\begin{align}
  &\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) \notag\\ &=
  \left(\frac{1}{\gamma}-\frac{1}{2}\right) +\frac{\ii}{2}\epsilon q
  \hat{k} + \frac{1}{2}\left(2M+1\right)\left(A^2 + 4C\right)^{1/2}\label{adia_disp}. 
\end{align}
This gives the eigenfrequency $\hat{\sigma}$ and completes the solution
description.

@@

>>>>

\subsubsection{Explicit solutions in the thin-disk limit}
We can construct solutions in the thin-disk limit, in which 
$\ln(\rho/\rho_0) \simeq -\zhat^2/2$. Eq. \ref{adia_iso3} becomes   
\begin{align}
  0 = \delta v_z^{\prime\prime} - \zhat A \delta v_z^\prime + \left(B
    - C \zhat^2\right)\delta v_z, \label{adia_iso4} 
\end{align}
with
\begin{align}
  &A \equiv 1 + \ii \epsilon q \hat{k},\\
  &B \equiv \hat{\sigma}^2\left(\frac{1}{\gamma} + \hat{k}^2\right) -
  \left(\frac{1}{\gamma} + \ii \hat{k} \epsilon q\right)\\
  &C \equiv \frac{\left(\gamma-1\right)}{\gamma}\hat{k}^2\left(1 - \frac{\ii
      \epsilon q}{\hat{k}}\right).\label{adia_thin}
\end{align}
Let
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\alpha\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where $\alpha$ is a constant to be chosen for convenience. Inserting
Eq. \ref{adia_ansatz} into Eq. \ref{adia_thin} gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime + \left(B +
    \alpha\right)g
  +\left(\alpha^2 - \alpha A - C\right)\zhat^2 g.
\end{align}
We choose $\alpha$ to make the coefficient of $\zhat^2g$
vanish, and impose the vertical kinetic energy density
$\rho|\delta v_z|^2$ to remain finite. Then assuming $g(\zhat)$ is a
polynomial, we require  
\begin{align}
  \real\alpha < \frac{1}{2}. 
\end{align}
We therefore choose 
\begin{align}
  \alpha = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right). 
\end{align}

We now have an equation in the same form as Eq. \ref{iso_ode3}:
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
% We again seek power-series solutions 
% \begin{align}
%   g(\zhat) = \sum_{m=0}^\infty b_m \zhat^m,
% \end{align}
% which requires
% \begin{align}
%   (m+2)(m+1)b_{m+2} + \left[\left(2\alpha - A\right)m + B + \alpha\right]=0.
% \end{align}
which requires
\begin{align}
  B + \alpha = M\left(A-2\alpha\right).\label{adia_disp0}
\end{align}
Inserting the definition of $B$ we get
\begin{align}
  &\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) \notag\\ &=
  \left(\frac{1}{\gamma}-\frac{1}{2}\right) +\frac{\ii}{2}\epsilon q
  \hat{k} + \frac{1}{2}\left(2M+1\right)\left(A^2 + 4C\right)^{1/2}\label{adia_disp}. 
\end{align}
This gives the eigenfrequency $\hat{\sigma}$ and completes the solution
description.

@@

\subsection{Perturbations with thermal relaxation}
Having examined the extreme limits of isothermal ($t_c\to0$) and
adiabatic ($t_c\to\infty$) perturbations, we now generalize further to consider
perturbations with intermediate thermodynamic responses with general
values of $t_c$.

In this section we will pursue explicit solutions in the low-frequency,
thin-disk, nearly-Keplerian approximations as used in the development
for adiabatic solutions. The pair of ODEs for $W$ and $Q$
(Eq. \ref{ode_w}---\ref{ode_Q}) with these approximations and 
written with dimensionless variables, are
\begin{align}
  &\hat{\sigma}^2\left(Q +  \hat{k}^2W\right) \notag\\ 
  &= \hat{z} \left(1 +
    \ii\epsilon q \hat{k} \right)\left[W^\prime - \zhat \left(W -
      Q\right)\right] - \left[W^\prime - \zhat \left(W -
      Q\right)\right]^\prime\label{nearly_iso1},\\
&\beta \hat{\sigma}^2\left(W - \gamma Q\right) + \ii\hat{\sigma}
    \left(W-Q\right)\notag\\
   & = -\beta\zhat\left(\gamma-1\right) \left[W^\prime - \zhat \left(W -
      Q\right)\right]\label{nearly_iso2},
\end{align}
where we recall $\beta = t_c\Omega_k$ is the dimensionless cooling
time. It is evident that for $\beta\equiv 0$, we recover
Eq. \ref{iso_ode3}. In addition, from the expression of the vertical
velocity perturbation (Eq. \ref{lin_vz}) we have
\begin{align}
  W^\prime - \zhat \left(W -
      Q\right) = \ii\hat{\sigma}c_s\delta v_z. \label{nearly_vz}
\end{align}

\subsubsection{Perturbative effect of introducing a small but finite
  thermal relaxation time}

It is instructive to first ask: how
do the eigenfrequencies and eigenfunctions change when we change
$t_c$ from zero to a small but finite value? For sufficiently small
$t_c$, for which the perturbations may be considered
nearly-isothermal, we expect the new solutions to only differ slighly
from the explicit isothermal solutions developed in \S\ref{iso_poly}.

%is valid for
%non-isothermal perturbations $\gamma>1$ if the thermal relaxation
%(cooling) timescale $t_c\equiv0$.

% Then we may linearize Eq. \ref{nearly_iso1}---\ref{nearly_iso2}

To determine the (assumed) small effect of introducing a short thermal
relaxation, we linearize Eq. \ref{nearly_iso1}---\ref{nearly_iso2} about a known
solution for $\beta\equiv0$. Let 
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, W\to W+\delta W,\, Q \to Q+\delta
  Q,\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align} 
where $W=Q$ is the polynomial solution described in \S\ref{iso_poly}
with $\hat{\sigma}^2$ given by Eq. \ref{sig2_iso}. Inserting
Eq. \ref{nearly_iso_pert} into
Eq. \ref{nearly_iso1}---\ref{nearly_iso2} and keeping only first order
terms, we obtain
\begin{align}
  &2\hat{\sigma}\delta\hat{\sigma}\left(1+\hat{k}^2\right)W +
  \hat{\sigma}^2 \left(\delta Q + \hat{k}^2\delta W\right)\notag\\
  &= \zhat \left(1+\ii\epsilon q \hat{k}\right)\left[\delta W^\prime -
    \zhat \left(\delta W - 
      \delta Q\right)\right] - \delta W^{\prime\prime}\notag\\
  &\phantom{=} + \left[\zhat \left(\delta W -
     \delta Q\right)\right]^\prime,\label{nearly_iso3}\\
&\ii\hat{\sigma} \left(\delta W - \delta Q\right) =
\delta\beta\left(\gamma-1\right)\left(\hat{\sigma}^2W - \hat{z}W^\prime\right).\label{nearly_iso4}
\end{align}
This is a pair of inhomogeneous ODEs with a forcing by $W$. The
homogeneous problem is, of course, just the ODE for $\beta=0$ solved
in \S\ref{iso_poly}. Hence we only seek the particular solution.

We could proceed by considering the entire set of polynomial
solutions for $W$, but it is simplest to look at the fundamental
mode. Thus we consider the `basic state' 
\begin{align*}
  W = \hat{z},\quad \hat{\sigma}^2\left(1+\hat{k}^2\right) = 1 +
  \ii\epsilon q \hat{k},
\end{align*}
then
\begin{align}
  \delta W = b \zhat^3,
\end{align}
where $b$ is a constant to be determined. Eliminating $\delta Q$ from
Eq. \ref{nearly_iso3}---\ref{nearly_iso4}, inserting the above
expressions for $W$ and $\delta W$ and balancing coefficients of
$\zhat$ and $\zhat^3$, we find
\begin{align}
 & 2\left(1+\ii\epsilon q \hat{k}\right) \delta\hat{\sigma} \notag\\& = -6b\hat{\sigma} -
2\ii \delta \beta\left(\gamma-1\right)\left(\hat{\sigma}^2-1\right) - \ii
  \hat{\sigma}^2\delta\beta\left(\gamma-1\right),\\
& 0=  2b\hat{\sigma}+\ii\delta\beta
\left(\gamma-1\right)\left(\hat{\sigma}^2 - 1\right),
\end{align}
where the expression for $\hat{\sigma}^2$ was used. Solving for
$\delta\hat{\sigma}$ gives 
\begin{align}\label{nearly_iso_dsig}
  \delta \hat{\sigma} =
  -\frac{\ii\left(\gamma-1\right)\hat{k}^2\left(\ii\epsilon
    q - \hat{k}\right)^2}{2\left(1+\ii\epsilon q \hat{k}\right)\left(1+\hat{k}^2\right)^2}\delta\beta.
\end{align}
The imaginary part of Eq. \ref{nearly_iso_dsig} is
\begin{align}
   \delta\hat{\nu} =
  -\frac{\left(\gamma-1\right)\hat{k}^2 \left(\hat{k}^2 -
      2\epsilon^2q^2\hat{k}^2 - \epsilon^2q^2\right)}{2\left(1+\epsilon^2 q^2
        \hat{k}^2\right)\left(1+\hat{k}^2\right)^2}\delta\beta.  
\end{align}
Since $\epsilon \ll 1$, introducing finite cooling $\delta\beta>0$
implies $\delta\hat{\nu} < 0$, i.e. stabilization. Interestingly, for both
$\hat{k}\to0$ and $\hat{k}\to\infty$ we find $\delta\hat{\nu}\to0$, so
stabilization by finite cooling is ineffective at very large or very
small scales. We plot Eq. \ref{nearly_iso_dsig} in Fig. \ref{domegadbeta} with
$\epsilon=0.05$, $q=-1.0$ and $\gamma=1.4$. For this disk model the
optimum wavenumber for stabilization is $\hat{k}\sim 5$---$6$.

% This contrasts to the case without vertical shear
% ($q=0$), in which $\p\hat{\sigma}/\p\beta\to\mathrm{constant}$ as 
% $\hat{k}\to\infty$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/domegadbeta}
  \caption{Rate of change of the fundamental VSI eigenfrequency
    $\sigma$ with respect to the introduction of a small but finite
    cooling time. The plotted quantity is $\p\hat{\sigma}/\p\beta$ at $\beta=0$, evaluated using
    Eq. \ref{nearly_iso_dsig} with 
    disk parameters $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$. The real
    (imaginary) part of $\p\hat{\sigma}/\p\beta$ is shown as the solid
    (dotted) lines. 
    \label{domegadbeta}}  
\end{figure}

One effect of introducing finite thermal relaxation is buoyancy. % We
% show in the next section that buoyancy strongly stabilizes the
% VSI.
The solution $\delta W\propto\zhat^3$ then makes sense because it
is most significant for large $\zhat$, i.e. away from the midplane
where the effect of bouyancy first appears as $t_c$ is increased.

\subsubsection{General explicit solutions with thermal relaxation}
To develop explicit solutions of
Eq. \ref{nearly_iso1}---\ref{nearly_vz}, we eliminate variables in
favor of $\delta v_z$ to obtain
\begin{align}
  0 = \delta v_z^{\prime\prime} - z \widetilde{A}\delta v_z^\prime +
  (\widetilde{B} - \widetilde{C}\zhat^2)\delta v_z = 0,\label{nearly_iso_explicit}
\end{align}
with
\begin{align}
  &\widetilde{A} \equiv 1 + \ii \epsilon q \hat{k},\\
  &\widetilde{B} \equiv \hat{\sigma}^2\left(\chi + \hat{k}^2\right) -
  \left(\chi + \ii \epsilon q \hat{k}\right),\\
  &\widetilde{C} \equiv \left(1-\chi\right)\left(\hat{k}^2 - \ii
      \epsilon q\hat{k}\right),
\end{align}
where
\begin{align}
  \chi \equiv \frac{1-\ii\hat{\sigma}\beta}{1-\ii\hat{\sigma}\beta\gamma}.
\end{align}

The governing equation Eq. \ref{nearly_iso_explicit} has the same form
as in the adiabatic case (Eq. \ref{adia_iso4}), so we can apply the
same solution method. However, the coefficients
$\widetilde{B}$ and $\widetilde{C}$ now have a more complicated
dependence on $\hat{\sigma}$.  The adiabatic dispersion relation,
Eq. \ref{adia_disp0}, generalizes to a $6^\mathrm{th}$ order
polynomial in $\hat{\sigma}$:
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
Eq. \ref{relax_disp} contrasts to Eq. \ref{adia_disp}, which is
quadratic in $\hat{\sigma}$ for adiabatic perturbations.

It is simplest to solve Eq. \ref{relax_disp} numerically. In
Fig. \ref{relax_disp_fig} we plot the growth rate of the fundamental
($M=0$) VSI mode as a function of $k_xH_\mathrm{iso}$ for a range of
dimensionless thermal relaxation timescales $\beta$, in a disk with
$\epsilon=0.05$, $q=-1$ and $\gamma=1.4$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/rate_theory_grow_relax}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/rate_theory_grow_relax2}
  \caption{Growth rate of the fundamental VSI mode in a disk with
    dimensionless thermal relaxation timescales
    $\beta\in[10^{-3},1]$ (top) and $\beta\in[1,10^3]$ (bottom).
Other disk disk parameters are
    $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$. The growth rates are
    calculated from the dispersion relation Eq. \ref{relax_disp}. 
    \label{relax_disp_fig}}  
\end{figure}

%ate_theory, eps=0.05, smallq=-1d0, gmma=1.4, bmax=1.0, ncases=11, mmode=0.0, root=1, legend=[0.102,0.14,5.4,0.355], yrange=[0,6]

### 2014-12-27 linear_iso.tex
updated figures log axis
extended analytic discussion to thermal relaxation cases

\section{Analytical discussion of vertically isothermal disks} 
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the shearing box. We
examine separately the limiting case of isothermal perturbations
($\gamma\equiv1$) and adiabatic perturbations ($\gamma\neq 1$ and
$t_c\to\infty$).   
\subsection{Isothermal perturbations}
The linear problem simplifies considerably for the case
$\Gamma=\gamma=1$. Then 
\begin{align}
  Q=W. 
\end{align}
Inserting this into Eq. \ref{ode_w} gives a single equation for the linear stability
problem for isothermal perturbations in a vertically isothermal disk, 
\begin{align}\label{iso_ode}
  \frac{d^2W}{dz^2} + \left(\frac{d\ln{\rho}}{dz} - \frac{\ii k_x
      r}{D}\frac{d\Omega^2}{dz}\right) \frac{dW}{dz} +
  \sigma^2\left(\frac{1}{c_s^2} + \frac{k_x^2}{D}\right)W=0, 
\end{align} 
subject to appropriate boundary conditions. We remark that Eq
.\ref{iso_ode} is also applicable for    
$\gamma=\Gamma\neq 1$ in the limit $t_c\to\infty$ (i.e. neutrally
stratified adiabatic disks), since in that case  
$Q\simeq W$ as well. However, the following discussion is only
valid for $\gamma=\Gamma=1$ because $c_s$ is taken to be a constant.

>>>>

\section{Analytical discussion of vertically isothermal disks} 
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the shearing box. We
examine separately the limiting case of isothermal perturbations
($\gamma\equiv1$ and any value of $t_c$, or any value of $\gamma$ with
$t_c = 0$) and adiabatic perturbations ($\gamma\neq 1$ and $t_c\to\infty$).   
\subsection{Isothermal perturbations}\label{iso_discuss}
The linear problem simplifies considerably for the case
$\Gamma=\gamma=1$.  Then 
\begin{align}
  Q=W. 
\end{align}
Inserting this into Eq. \ref{ode_w} gives a single equation for the linear stability
problem for isothermal perturbations in a vertically isothermal disk, 
\begin{align}\label{iso_ode}
  \frac{d^2W}{dz^2} + \left(\frac{d\ln{\rho}}{dz} - \frac{\ii k_x
      r}{D}\frac{d\Omega^2}{dz}\right) \frac{dW}{dz} +
  \sigma^2\left(\frac{1}{c_s^2} + \frac{k_x^2}{D}\right)W=0, 
\end{align} 
subject to appropriate boundary conditions. Eq. \ref{iso_ode} also
applies to $\gamma>1$ for $t_c\to0$.

We remark that Eq.\ref{iso_ode} is also applicable for    
$\gamma=\Gamma\neq 1$ in the limit $t_c\to\infty$ (i.e. neutrally
stratified adiabatic disks), since in that case  
$Q\simeq W$ as well. However, the following discussion is only
valid for $\gamma=\Gamma=1$ because $c_s$ is taken to be a constant.
